---
title: "Modelling the Impact of Macroeconomic Factors on Insurance Premiums in Kenya"
subtitle: "Chapter 5 — Summary, Conclusions and Recommendations"
author:
 - "Whitney Kemuma"
 - "Shebunah Nyairabu"
date: |
  2025
  **Supervisor:** Dr Cynthia Mwende
  Department of Mathematics and Actuarial Science, Kisii University
output:
  pdf_document:
    latex_engine: xelatex
    toc: true
    toc_depth: 3
    number_sections: true
    fig_caption: true
geometry: "left=2.5cm, right=2.5cm, top=2.5cm, bottom=2.5cm"
header-includes:
  - \usepackage{booktabs}
  - \usepackage{float}
  - \usepackage{amsmath}
  - \usepackage{longtable}
  - \usepackage{array}
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo      = TRUE,
  message   = FALSE,
  warning   = FALSE,
  fig.pos   = "H",
  fig.align = "center",
  out.width = "90%",
  fig.width = 8,
  fig.height = 4.5
)
```

\newpage

# Chapter 5: Summary, Conclusions and Recommendations

## 5.1 Introduction

This chapter synthesises the findings from Chapters 1 through 4, answers
the research questions and hypotheses posed in Chapter 0, and offers
recommendations for actuarial practice, regulatory policy, and future
research. The study modelled the impact of macroeconomic factors on
comprehensive car insurance premiums in Kenya over the period 2015 to 2024,
using ARIMAX time series modelling and Multiple Linear Regression.

---

## 5.2 Load Libraries and Refit Models
```{r load-and-refit}
library(ggplot2)
library(dplyr)
library(knitr)
library(kableExtra)
library(forecast)
library(tseries)
library(car)
library(scales)

# Load data
df <- read.csv("../data/insurance_data.csv",
               stringsAsFactors = FALSE)
names(df) <- c("Year","Premium","Inflation",
               "GDP_Growth","Interest_Rate")

# Refit ARIMAX
premium_ts  <- ts(df$Premium, start = 2015, frequency = 1)
xreg_matrix <- cbind(
  Inflation = df$Inflation,
  GDP       = df$GDP_Growth,
  Interest  = df$Interest_Rate
)
arimax_model <- auto.arima(premium_ts, xreg = xreg_matrix,
                            stepwise = FALSE,
                            approximation = FALSE)

# Refit MLR
mlr_model <- lm(Premium ~ Inflation + GDP_Growth + Interest_Rate,
                data = df)
mlr_sum   <- summary(mlr_model)

cat("Models refitted.\n")
```

---

## 5.3 Summary of Findings

### 5.3.1 Research Question 1 — Trends and Patterns

Insurance premiums in Kenya followed a U-shaped trajectory between 2015
and 2024, declining from KES 32,000 in 2015 to KES 30,000 in 2018 before
recovering to KES 37,500 by 2024 — a net increase of 17.2% over the study
period. The initial decline coincided with falling interest rates and
moderate inflation, while the post-2019 upward trend aligns with rising
claim costs and inflationary pressure.

Interest rates declined monotonically by 5 percentage points over the period,
representing a structural shift in the investment environment for Kenyan
insurers. Inflation fluctuated around a mean of
`r round(mean(df$Inflation), 2)`%, with notable peaks in 2017 and 2022-2023.
GDP growth was broadly positive except for the COVID-19 contraction of -0.27%
in 2020.

### 5.3.2 Research Question 2 — ARIMAX Model Performance
```{r arimax-summary-table}
acc  <- accuracy(arimax_model)
lb   <- Box.test(residuals(arimax_model), lag = 3,
                 type = "Ljung-Box")

arimax_df <- data.frame(
  Metric = c("ARIMA Order", "AIC", "BIC",
             "MAPE (%)", "RMSE (KES)",
             "Ljung-Box p-value",
             "Residual white noise?"),
  Value  = c(
    paste0("ARIMA(", arimax_model$arma[1], ",",
           arimax_model$arma[6], ",",
           arimax_model$arma[2], ")"),
    round(arimax_model$aic,  2),
    round(arimax_model$bic,  2),
    round(acc[1,"MAPE"],     2),
    round(acc[1,"RMSE"],     2),
    round(lb$p.value,        4),
    ifelse(lb$p.value > 0.05, "Yes", "No")
  )
)

kable(arimax_df,
      caption  = "ARIMAX Model Performance Summary",
      booktabs = TRUE,
      align    = c("l","r")) %>%
  kable_styling(
    latex_options = c("striped", "hold_position"),
    font_size     = 10,
    full_width    = FALSE
  )
```

The ARIMAX model provided a good fit to the historical data with a MAPE
below 5%, and the Ljung-Box test confirmed that residuals behaved as
white noise, indicating no systematic patterns were left unexplained.

### 5.3.3 Research Question 3 — Hypothesis Test Outcomes
```{r hyp-summary-table}
coef_mat <- mlr_sum$coefficients
p_inf    <- coef_mat["Inflation",     4]
p_gdp    <- coef_mat["GDP_Growth",    4]
p_int    <- coef_mat["Interest_Rate", 4]
b_inf    <- coef_mat["Inflation",     1]
b_gdp    <- coef_mat["GDP_Growth",    1]
b_int    <- coef_mat["Interest_Rate", 1]

f_stat   <- mlr_sum$fstatistic
f_pval   <- pf(f_stat[1], f_stat[2], f_stat[3],
               lower.tail = FALSE)

hyp_df <- data.frame(
  Hypothesis = c(
    "H1: Inflation effect on premiums",
    "H2: GDP growth effect on premiums",
    "H3: Interest rate effect on premiums",
    "H4: Overall model significance"
  ),
  Coefficient = c(
    round(b_inf, 2),
    round(b_gdp, 2),
    round(b_int, 2),
    NA
  ),
  P_Value = c(
    round(p_inf,   4),
    round(p_gdp,   4),
    round(p_int,   4),
    round(f_pval,  4)
  ),
  Decision = c(
    ifelse(p_inf  < 0.05, "Reject H0", "Fail to Reject H0"),
    ifelse(p_gdp  < 0.05, "Reject H0", "Fail to Reject H0"),
    ifelse(p_int  < 0.05, "Reject H0", "Fail to Reject H0"),
    ifelse(f_pval < 0.05, "Reject H0", "Fail to Reject H0")
  ),
  Direction = c(
    ifelse(b_inf > 0, "Positive (expected)", "Negative"),
    ifelse(b_gdp > 0, "Positive (expected)", "Negative"),
    ifelse(b_int < 0, "Negative (expected)", "Positive"),
    "—"
  )
)

kable(hyp_df,
      col.names = c("Hypothesis", "Coefficient",
                    "p-value", "Decision", "Direction"),
      caption   = "Hypothesis Test Results Summary (alpha = 0.05)",
      booktabs  = TRUE,
      align     = c("l","r","r","l","l")) %>%
  kable_styling(
    latex_options = c("striped", "hold_position", "scale_down"),
    font_size     = 10,
    full_width    = FALSE
  ) %>%
  footnote(
    general = paste(
      "All tests conducted at the 5% significance level.",
      "Coefficient shown is the MLR estimate.",
      "For H4, the test statistic is the F-statistic."
    ),
    footnote_as_chunk = TRUE,
    threeparttable    = TRUE
  )
```

### 5.3.4 Research Question 4 — Premium Forecasts 2025-2027
```{r forecast-summary-table}
xreg_base <- matrix(
  c(3.50, 5.60, 9.00,
    3.50, 5.60, 9.00,
    3.50, 5.60, 9.00),
  nrow = 3, ncol = 3, byrow = TRUE,
  dimnames = list(NULL,
                  c("Inflation","GDP","Interest"))
)

fc_base <- forecast(arimax_model,
                    xreg = xreg_base, h = 3)

fc_df <- data.frame(
  Year     = 2025:2027,
  Forecast = round(as.numeric(fc_base$mean), 0),
  Lo_80    = round(as.numeric(fc_base$lower[,1]), 0),
  Hi_80    = round(as.numeric(fc_base$upper[,1]), 0),
  Lo_95    = round(as.numeric(fc_base$lower[,2]), 0),
  Hi_95    = round(as.numeric(fc_base$upper[,2]), 0)
)

kable(fc_df,
      col.names = c("Year", "Point Forecast (KES)",
                    "80% CI Lower", "80% CI Upper",
                    "95% CI Lower", "95% CI Upper"),
      caption   = "Baseline Premium Forecasts 2025-2027 with Confidence Intervals",
      booktabs  = TRUE,
      align     = c("c","r","r","r","r","r"),
      format.args = list(big.mark = ",")) %>%
  kable_styling(
    latex_options = c("striped", "hold_position", "scale_down"),
    font_size     = 10,
    full_width    = FALSE
  ) %>%
  footnote(
    general = "Baseline assumes 2024 macroeconomic levels held constant.
    CI = confidence interval.",
    footnote_as_chunk = TRUE,
    threeparttable    = TRUE
  )
```

---

## 5.4 Conclusions

Based on the empirical analysis conducted in this study, the following
conclusions are drawn:

**Conclusion 1 — Interest rate is the dominant macroeconomic predictor.**
The interest rate was the only variable achieving statistical significance
at the 5% level in the MLR model
(β = `r round(b_int, 2)`, p = `r round(p_int, 4)`).
The strong negative correlation (r = `r round(cor(df$Premium, df$Interest_Rate), 3)`)
and significant coefficient confirm that the decade long decline in the CBK
benchmark rate has been a primary driver of upward premium pressure in Kenya.
This is consistent with economic theory, as investment income from insurer
asset portfolios falls with declining rates, premium income must compensate.

**Conclusion 2 — Inflation and GDP are directionally consistent but statistically weak.**
Both inflation and GDP growth showed the theoretically expected signs —
positive effects on premiums, but neither achieved significance at the 5%
level over this study period. The small sample size (n = 10) substantially
limits statistical power, and these variables may become significant with
a longer data series.

**Conclusion 3 — The ARIMAX model provides an adequate fit.**
The model achieved a MAPE of `r round(accuracy(arimax_model)[1,"MAPE"], 2)`%
on training data, and residual diagnostics confirmed white noise behaviour
(Ljung-Box p = `r round(Box.test(residuals(arimax_model), lag=3, type="Ljung-Box")$p.value, 4)`).
The actual vs fitted plot showed close alignment from 2018 onward,
with modest early-period deviation.

**Conclusion 4 — Premiums are projected to continue rising.**
Under baseline macroeconomic assumptions, the ARIMAX forecast projects
premiums reaching KES `r fc_df$Forecast[1]` in 2025,
KES `r fc_df$Forecast[2]` in 2026, and
KES `r fc_df$Forecast[3]` in 2027. The widening confidence intervals
reflect increasing forecast uncertainty over the horizon and underscore
the importance of monitoring macroeconomic developments.

---

## 5.5 Recommendations

### 5.5.1 For Insurance Companies

**Embed interest rate monitoring into pricing cycles.** Given the dominant
role of interest rates, insurers should integrate CBK rate decisions and
interest rate forecasts directly into actuarial pricing reviews. A 1
percentage point change in the benchmark rate is associated with a
KES `r round(abs(b_int), 0)` change in annual premiums according to the
MLR model.

**Develop inflation-indexed pricing reserves.** Although inflation did not
achieve statistical significance in this study, its directional positive
effect is consistent with theory. Insurers should maintain inflation
sensitivity buffers, particularly for long tail lines where claims are
settled years after premium collection.

**Use scenario forecasting for strategic planning.** The three scenario
forecast framework developed in Chapter 3 provides a practical tool for
stress testing pricing assumptions against optimistic and pessimistic
macroeconomic outcomes.

### 5.5.2 For Policymakers and Regulators

**Macroeconomic stability directly supports insurance market stability.**
The significant link between interest rates and premiums means that CBK
monetary policy has direct implications for insurance affordability.
Regulators should account for interest rate environments when assessing
whether premium adjustments are economically justified.

**Promote transparency in premium adjustment reporting.** Regulatory
frameworks should require insurers to document the macroeconomic
assumptions underpinning premium changes, enabling regulators to
verify that adjustments reflect genuine economic drivers rather than
opportunistic pricing.

### 5.5.3 For Future Researchers

**Extend the dataset.** The most significant limitation of this study is
the small sample size (n = 10). Extending the analysis to quarterly data
or a longer annual series would substantially improve statistical power
and enable more robust inference.

**Include additional macroeconomic variables.** Unemployment, exchange
rates, and consumer confidence indices were not included due to data
availability constraints. Their inclusion may strengthen the predictive
power of the model.

**Disaggregate by insurance product line.** This study used a single
insurance product (comprehensive motor). Different lines — life, health,
property - may respond differently to macroeconomic conditions,
offering richer insights for insurers operating across multiple segments.

**Apply machine learning approaches.** For longer datasets, gradient
boosting and neural network time series models (LSTM) could be compared
against the ARIMAX and MLR baselines to assess whether non-linear methods
offer predictive improvements in the Kenyan insurance context.

---

## 5.6 Limitations
```{r limitations-table, echo=FALSE}
lim_df <- data.frame(
  Limitation = c(
    "Small sample size (n = 10)",
    "Annual data frequency",
    "Single insurance product",
    "Three predictors only",
    "Interest rate assumed from trend",
    "No structural break modelling"
  ),
  Impact = c(
    "Low statistical power — limits significance testing",
    "Cannot capture intra-year seasonality",
    "Findings may not generalise to life or health insurance",
    "Omitted variables may bias coefficient estimates",
    "Assumed values introduce measurement uncertainty",
    "COVID-19 shock treated as a data point, not a structural break"
  ),
  Mitigation = c(
    "LOO cross-validation used; results interpreted with appropriate caution",
    "Annual data is appropriate given annual premium setting cycles",
    "Explicitly acknowledged in scope (Chapter 0, Section 1.4)",
    "Correlation matrix and VIF confirm no multicollinearity among included variables",
    "Sensitivity analysis across scenarios conducted in Chapter 3",
    "2020 flagged throughout; residual plots checked for outlier influence"
  )
)

kable(lim_df,
      col.names = c("Limitation", "Impact", "Mitigation"),
      caption   = "Study Limitations, Impacts and Mitigations",
      booktabs  = TRUE) %>%
  kable_styling(
    latex_options = c("striped", "hold_position", "scale_down"),
    font_size     = 9,
    full_width    = FALSE
  ) %>%
  column_spec(1, width = "3.5cm") %>%
  column_spec(2, width = "4.5cm") %>%
  column_spec(3, width = "5cm")
```

---

## 5.7 Final Project Summary
```{r final-summary, echo=FALSE}
cat("=======================================================\n")
cat("  PROJECT COMPLETE — FULL STUDY SUMMARY\n")
cat("=======================================================\n\n")
cat("Title   : Modelling the Impact of Macroeconomic Factors\n")
cat("          on Insurance Premiums Using Time Series\n")
cat("          Analysis and Multiple Linear Regression\n")
cat("Author  : Whitney Kemuma\n")
cat("Period  : 2015-2024 (Kenya)\n\n")
cat("Models  :\n")
cat("  ARIMAX : ARIMA(",
    arimax_model$arma[1], ",",
    arimax_model$arma[6], ",",
    arimax_model$arma[2], ") with 3 exogenous regressors\n")
cat("  MLR    : Premium ~ Inflation + GDP + Interest Rate\n\n")
cat("Key results:\n")
cat("  R-squared (MLR)  :", round(mlr_sum$r.squared, 4), "\n")
cat("  MAPE (ARIMAX)    :",
    round(accuracy(arimax_model)[1,"MAPE"], 2), "%\n")
cat("  Significant pred : Interest Rate (p =",
    round(p_int, 4), ")\n\n")
cat("Forecasts (baseline, 2024 macro levels):\n")
cat("  2025 :", fc_df$Forecast[1], "KES\n")
cat("  2026 :", fc_df$Forecast[2], "KES\n")
cat("  2027 :", fc_df$Forecast[3], "KES\n\n")
cat("Chapters: 0 (Foundation) | 1 (Data) | 2 (EDA)\n")
cat("          3 (ARIMAX) | 4 (MLR) | 5 (Conclusions)\n\n")
cat("  Project submission ready.\n")
cat("=======================================================\n")
```

---

# References

- Aziz, A.M.A.K., & Aziz, M.A.H.M.A.K. (2020). Macroeconomic variables
  and insurance premiums in Malaysia. *Google Scholar.*

- Box, G.E.P., & Jenkins, G.M. (1976). *Time Series Analysis: Forecasting
  and Control.* Holden-Day.

- Central Bank of Kenya (2023). *Monetary Policy Statement — June 2023.*
  Retrieved from https://www.centralbank.go.ke

- EIOPA (2022). *Insurance and Inflation.*
  Retrieved from https://www.eiopa.europa.eu

- Frees, E.W., Derrig, R.A., & Meyers, G. (2014). *Predictive Modeling
  Applications in Actuarial Science.* Cambridge University Press.

- Insurance Regulatory Authority (IRA) (2023). *Annual Insurance Industry
  Report.* Retrieved from https://www.ira.go.ke

- Kenya National Bureau of Statistics (KNBS) (2023). *Economic Survey 2023.*
  Retrieved from https://www.knbs.or.ke

- Kimani, J. (2021). Inflation and financial performance of Kenyan insurance
  companies. *eRepository.*

- Mwangi, M. (2017). Effect of macroeconomic variables on financial
  performance of insurance firms in Kenya. *eRepository.*

- Njeri, W. (2016). Interest rates and financial performance of insurance
  firms in Kenya. *eRepository.*

- Nyambura, L. (2022). Time series analysis of insurance claims in Kenya.
  *Kenyatta University.*

- Outreville, J.F. (2020). The relationship between insurance and economic
  development. *SpringerLink.*