---
title: "Modelling the Impact of Macroeconomic Factors on Insurance Premiums in Kenya"
subtitle: "Chapter 2 — Exploratory Data Analysis"
author:
 - "Whitney Kemuma"
 - "Shebunah Nyairabu"
date: |
  2025
  **Supervisor:** Dr Cynthia Mwende
  Department of Mathematics and Actuarial Science, Kisii University
output:
  pdf_document:
    latex_engine: xelatex
    toc: true
    toc_depth: 3
    number_sections: true
    fig_caption: true
    keep_tex: false
geometry: "left=2.5cm, right=2.5cm, top=2.5cm, bottom=2.5cm"
header-includes:
  - \usepackage{booktabs}
  - \usepackage{float}
  - \usepackage{amsmath}
  - \usepackage{graphicx}
  - \usepackage{longtable}
  - \usepackage{array}
  - \usepackage{pdflscape}
  - \setlength{\LTcapwidth}{\textwidth}
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo      = TRUE,
  message   = FALSE,
  warning   = FALSE,
  fig.pos   = "H",
  fig.align = "center",
  out.width = "90%",
  fig.width = 8,
  fig.height = 4.5
)
```

\newpage

# Chapter 2: Exploratory Data Analysis

## 2.1 Introduction

Exploratory Data Analysis (EDA) is conducted before modelling to understand
the structure, trends, and relationships in the data. The findings in this
chapter directly motivate the modelling choices in Chapters 3 and 4,
particularly the selection of ARIMAX for time series modelling and the
assessment of which macroeconomic variables carry the most predictive signal.

This chapter covers:

| Section | Activity |
|---------|----------|
| 2.2 | Reload data and libraries |
| 2.3 | Individual trend plots for all four variables |
| 2.4 | Combined multi-panel trend overview |
| 2.5 | Correlation analysis |
| 2.6 | Stationarity testing (ADF test) |
| 2.7 | Granger causality tests |
| 2.8 | EDA summary and modelling implications |

---

## 2.2 Reload Data and Libraries
```{r libraries-data}
library(ggplot2)
library(dplyr)
library(knitr)
library(kableExtra)
library(forecast)
library(tseries)
library(lmtest)
library(corrplot)
library(gridExtra)
library(scales)

df <- read.csv("../data/insurance_data.csv",
               stringsAsFactors = FALSE)
names(df) <- c("Year", "Premium", "Inflation",
               "GDP_Growth", "Interest_Rate")

premium_ts <- ts(df$Premium, start = 2015, frequency = 1)

cat("Data loaded. Observations:", nrow(df), "\n")
```

---

## 2.3 Individual Variable Trend Plots

### 2.3.1 Insurance Premium Trend
```{r plot-premium, fig.cap="Insurance Premium Trend (2015-2024)"}
ggplot(df, aes(x = Year, y = Premium)) +
  geom_line(color = "#2c7bb6", linewidth = 1.2) +
  geom_point(color = "#2c7bb6", size = 3) +
  geom_text(aes(label = paste0("KES\n",
            format(Premium, big.mark = ","))),
            vjust = -0.6, size = 2.8, color = "#1a1a2e") +
  labs(
    title   = "Figure 2.1 — Insurance Premium Over Time",
    x       = "Year",
    y       = "Annual Premium (KES)",
    caption = "Source: Insurance Regulatory Authority (IRA), 2015-2024"
  ) +
  scale_x_continuous(breaks = 2015:2024) +
  scale_y_continuous(labels = comma,
                     limits = c(27000, 41000)) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title   = element_text(face = "bold", size = 11),
    axis.text.x  = element_text(angle = 45, hjust = 1),
    plot.caption = element_text(size = 8, color = "grey50")
  )
```

**Interpretation:** Insurance premiums declined from KES 32,000 in 2015
to a minimum of KES 30,000 in 2018, before recovering and rising steadily
to KES 37,500 by 2024. The initial decline coincides with falling interest
rates and contained inflation, while the post 2019 increase aligns with
rising claim costs and inflationary pressures.

---

### 2.3.2 Inflation Rate Trend
```{r plot-inflation, fig.cap="Inflation Rate Trend (2015-2024)"}
ggplot(df, aes(x = Year, y = Inflation)) +
  geom_line(color = "#f4a261", linewidth = 1.2) +
  geom_point(color = "#f4a261", size = 3) +
  geom_text(aes(label = paste0(Inflation, "%")),
            vjust = -0.9, size = 3, color = "#1a1a2e") +
  geom_hline(yintercept = mean(df$Inflation),
             linetype = "dashed", color = "grey50",
             linewidth = 0.8) +
  annotate("text", x = 2015.3,
           y = mean(df$Inflation) + 0.25,
           label = paste0("Mean: ",
                          round(mean(df$Inflation), 2), "%"),
           size = 3, color = "grey40") +
  labs(
    title   = "Figure 2.2 — Inflation Rate Over Time",
    x       = "Year",
    y       = "Inflation Rate (%)",
    caption = "Source: Kenya National Bureau of Statistics, 2015-2024"
  ) +
  scale_x_continuous(breaks = 2015:2024) +
  scale_y_continuous(limits = c(2, 10)) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title   = element_text(face = "bold", size = 11),
    axis.text.x  = element_text(angle = 45, hjust = 1),
    plot.caption = element_text(size = 8, color = "grey50")
  )
```

**Interpretation:** Inflation fluctuated across the study period, peaking
at 8.01% in 2017 and reaching a decade low of 3.50% in 2024. The double
peak around 2017 and 2022-2023 reflects global commodity price shocks and
post COVID supply chain pressures respectively.

---

### 2.3.3 GDP Growth Rate Trend
```{r plot-gdp, fig.cap="GDP Growth Rate Trend (2015-2024)"}
ggplot(df, aes(x = Year, y = GDP_Growth)) +
  geom_col(aes(fill = GDP_Growth < 0),
           width = 0.6, alpha = 0.85) +
  scale_fill_manual(values = c("#2ecc71", "#e74c3c"),
                    guide  = "none") +
  geom_line(color = "#27ae60", linewidth = 1) +
  geom_point(color = "#27ae60", size = 3) +
  geom_text(aes(label = paste0(GDP_Growth, "%")),
            vjust = ifelse(df$GDP_Growth < 0, 1.8, -0.8),
            size  = 3, color = "#1a1a2e") +
  geom_hline(yintercept = 0, color = "black",
             linewidth = 0.8) +
  annotate("text", x = 2020, y = -1.5,
           label = "COVID-19", size = 3,
           color = "#e74c3c", fontface = "bold") +
  labs(
    title   = "Figure 2.3 — GDP Growth Rate Over Time",
    x       = "Year",
    y       = "GDP Growth Rate (%)",
    caption = "Source: World Bank / KNBS, 2015-2024. Red = negative growth."
  ) +
  scale_x_continuous(breaks = 2015:2024) +
  scale_y_continuous(limits = c(-2.5, 10)) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title   = element_text(face = "bold", size = 11),
    axis.text.x  = element_text(angle = 45, hjust = 1),
    plot.caption = element_text(size = 8, color = "grey50")
  )
```

**Interpretation:** GDP growth was broadly positive across the study period,
averaging `r round(mean(df$GDP_Growth), 2)`%. The 2020 contraction of
-0.27% represents the COVID-19 shock. The strong rebound to 7.59% in 2021
reflects post pandemic economic recovery.

---

### 2.3.4 Interest Rate Trend
```{r plot-interest, fig.cap="Interest Rate Trend (2015-2024)"}
ggplot(df, aes(x = Year, y = Interest_Rate)) +
  geom_ribbon(aes(ymin = min(Interest_Rate),
                  ymax = Interest_Rate),
              fill = "#e74c3c", alpha = 0.12) +
  geom_line(color = "#e74c3c", linewidth = 1.2) +
  geom_point(color = "#e74c3c", size = 3) +
  geom_text(aes(label = paste0(Interest_Rate, "%")),
            vjust = -0.9, size = 3, color = "#1a1a2e") +
  labs(
    title   = "Figure 2.4 — Interest Rate Over Time",
    x       = "Year",
    y       = "Interest Rate (%)",
    caption = "Source: Central Bank of Kenya (CBK), 2015-2024"
  ) +
  scale_x_continuous(breaks = 2015:2024) +
  scale_y_continuous(limits = c(7, 16)) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title   = element_text(face = "bold", size = 11),
    axis.text.x  = element_text(angle = 45, hjust = 1),
    plot.caption = element_text(size = 8, color = "grey50")
  )
```

**Interpretation:** The interest rate declined monotonically from 14%
in 2015 to 9% in 2024, a 5 percentage point reduction. This sustained
decline reflects the CBK's accommodative monetary policy and creates
pressure on insurer investment income, making premium income more critical.

---

## 2.4 Combined Multi-Panel Overview
```{r combined-plot, fig.cap="All Variables: Combined Trend Overview (2015-2024)", fig.height=8, fig.width=9}
p1 <- ggplot(df, aes(x = Year, y = Premium)) +
  geom_line(color = "#2c7bb6", linewidth = 1) +
  geom_point(color = "#2c7bb6", size = 2) +
  scale_y_continuous(labels = comma) +
  scale_x_continuous(breaks = seq(2015, 2024, 2)) +
  labs(title = "Insurance Premium (KES)", x = NULL, y = "KES") +
  theme_minimal(base_size = 10) +
  theme(plot.title = element_text(face = "bold", size = 10))

p2 <- ggplot(df, aes(x = Year, y = Inflation)) +
  geom_line(color = "#f4a261", linewidth = 1) +
  geom_point(color = "#f4a261", size = 2) +
  scale_x_continuous(breaks = seq(2015, 2024, 2)) +
  labs(title = "Inflation Rate (%)", x = NULL, y = "%") +
  theme_minimal(base_size = 10) +
  theme(plot.title = element_text(face = "bold", size = 10))

p3 <- ggplot(df, aes(x = Year, y = GDP_Growth)) +
  geom_col(aes(fill = GDP_Growth < 0), width = 0.6,
           alpha = 0.7) +
  scale_fill_manual(values = c("#2ecc71","#e74c3c"),
                    guide  = "none") +
  geom_line(color = "#27ae60", linewidth = 1) +
  scale_x_continuous(breaks = seq(2015, 2024, 2)) +
  geom_hline(yintercept = 0, linewidth = 0.5) +
  labs(title = "GDP Growth Rate (%)", x = NULL, y = "%") +
  theme_minimal(base_size = 10) +
  theme(plot.title = element_text(face = "bold", size = 10))

p4 <- ggplot(df, aes(x = Year, y = Interest_Rate)) +
  geom_line(color = "#e74c3c", linewidth = 1) +
  geom_point(color = "#e74c3c", size = 2) +
  scale_x_continuous(breaks = seq(2015, 2024, 2)) +
  labs(title = "Interest Rate (%)", x = "Year", y = "%") +
  theme_minimal(base_size = 10) +
  theme(plot.title = element_text(face = "bold", size = 10))

grid.arrange(p1, p2, p3, p4, ncol = 2,
  top = "Figure 2.5 — All Variables: Trend Overview (2015-2024)")
```

---

## 2.5 Correlation Analysis
```{r correlation-table}
cor_vars   <- df[, c("Premium", "Inflation",
                     "GDP_Growth", "Interest_Rate")]
cor_matrix <- cor(cor_vars, method = "pearson")

kable(round(cor_matrix, 4),
      caption  = "Pearson Correlation Matrix — All Variables",
      booktabs = TRUE,
      align    = "r") %>%
  kable_styling(
    latex_options = c("striped", "hold_position", "scale_down"),
    font_size     = 10,
    full_width    = FALSE
  )
```
```{r correlation-plot, fig.cap="Pearson Correlation Heatmap", fig.height=5, fig.width=5.5}
corrplot(cor_matrix,
         method      = "color",
         type        = "lower",
         addCoef.col = "black",
         tl.col      = "black",
         tl.cex      = 0.9,
         number.cex  = 0.9,
         col = colorRampPalette(
           c("#e74c3c", "white", "#2c7bb6"))(200),
         title = "Figure 2.6 — Pearson Correlation Heatmap",
         mar   = c(0, 0, 2, 0))
```

**Key observations:**

- **Premium vs Interest Rate:** r = `r round(cor_matrix["Premium","Interest_Rate"], 3)` — strong negative relationship. As rates fall, premiums rise.
- **Premium vs Inflation:** r = `r round(cor_matrix["Premium","Inflation"], 3)` — weak positive relationship.
- **Premium vs GDP Growth:** r = `r round(cor_matrix["Premium","GDP_Growth"], 3)` — weak positive relationship.
- No multicollinearity concerns between predictors.

---

## 2.6 Stationarity Testing — ADF Test

**Hypotheses (Chapter 0, Section 2.3.5):**

$$H_0: \text{Series has a unit root (non-stationary)}$$
$$H_1: \text{Series is stationary}$$
```{r adf-all}
vars <- c("Premium","Inflation","GDP_Growth","Interest_Rate")

adf_table <- data.frame(
  Variable  = vars,
  Statistic = NA_real_,
  P_Value   = NA_real_,
  Decision  = NA_character_,
  stringsAsFactors = FALSE
)

for (i in seq_along(vars)) {
  ts_var <- ts(df[[vars[i]]], start = 2015, frequency = 1)
  adf_r  <- adf.test(ts_var, alternative = "stationary")
  adf_table$Statistic[i] <- round(adf_r$statistic, 4)
  adf_table$P_Value[i]   <- round(adf_r$p.value,   4)
  adf_table$Decision[i]  <- ifelse(
    adf_r$p.value < 0.05, "Stationary", "Non-stationary"
  )
}

kable(adf_table,
      col.names = c("Variable", "ADF Statistic",
                    "p-value", "Decision (alpha=0.05)"),
      caption   = "ADF Stationarity Test Results — All Variables",
      booktabs  = TRUE,
      align     = c("l","r","r","l")) %>%
  kable_styling(
    latex_options = c("striped", "hold_position", "scale_down"),
    font_size     = 10,
    full_width    = FALSE
  ) %>%
  footnote(
    general = paste(
      "H0: Unit root present (non-stationary).",
      "Rejection at alpha = 0.05 indicates stationarity.",
      "Note: ADF test has limited power with n = 10 observations.",
      "auto.arima() in Chapter 3 handles differencing automatically."
    ),
    footnote_as_chunk = TRUE,
    threeparttable    = TRUE
  )
```

---

## 2.7 Granger Causality Tests

$$H_0: X \text{ does not Granger-cause Premium}$$
$$H_1: X \text{ does Granger-cause Premium}$$
```{r granger}
granger_vars  <- c("Inflation", "GDP_Growth", "Interest_Rate")
granger_table <- data.frame(
  Variable = granger_vars,
  F_Stat   = NA_real_,
  P_Value  = NA_real_,
  Decision = NA_character_,
  stringsAsFactors = FALSE
)

for (i in seq_along(granger_vars)) {
  pair <- cbind(df$Premium, df[[granger_vars[i]]])
  tryCatch({
    g <- grangertest(pair, order = 1)
    granger_table$F_Stat[i]  <- round(g$F[2], 4)
    granger_table$P_Value[i] <- round(g$`Pr(>F)`[2], 4)
    granger_table$Decision[i] <- ifelse(
      g$`Pr(>F)`[2] < 0.05,
      "Granger-causes",
      "Does not Granger-cause"
    )
  }, error = function(e) {
    granger_table$Decision[i] <<- "Test failed"
  })
}

kable(granger_table,
      col.names = c("Predictor", "F-Statistic",
                    "p-value", "Decision (alpha=0.05)"),
      caption   = "Granger Causality Test Results (Predictor to Premium, lag=1)",
      booktabs  = TRUE,
      align     = c("l","r","r","l")) %>%
  kable_styling(
    latex_options = c("striped", "hold_position", "scale_down"),
    font_size     = 10,
    full_width    = FALSE
  ) %>%
  footnote(
    general = paste(
      "H0: Predictor does not Granger-cause Premium.",
      "Lag = 1 due to small sample (n = 10).",
      "Granger causality is predictive, not structural causal."
    ),
    footnote_as_chunk = TRUE,
    threeparttable    = TRUE
  )
```

---

## 2.8 EDA Summary and Modelling Implications
```{r eda-summary, echo=FALSE}
cat("================================================\n")
cat("  CHAPTER 2 COMPLETE — EDA SUMMARY\n")
cat("================================================\n\n")
cat("FINDING 1 — Premium Trend\n")
cat("  Declined 2015-2018 then rose steadily to 2024.\n")
cat("  Total rise over study period: KES",
    max(df$Premium) - min(df$Premium), "\n\n")
cat("FINDING 2 — Inflation\n")
cat("  Fluctuating. Weak positive correlation with premiums.\n")
cat("  r =", round(cor_matrix["Premium","Inflation"], 3), "\n\n")
cat("FINDING 3 — GDP Growth\n")
cat("  COVID-19 shock in 2020. Weak positive correlation.\n")
cat("  r =", round(cor_matrix["Premium","GDP_Growth"], 3), "\n\n")
cat("FINDING 4 — Interest Rate\n")
cat("  Monotonic decline. Strong negative correlation.\n")
cat("  r =", round(cor_matrix["Premium","Interest_Rate"], 3), "\n\n")
cat("MODELLING IMPLICATIONS\n")
cat("  Interest rate is likely the dominant predictor.\n")
cat("  ADF results inform ARIMA order in Chapter 3.\n")
cat("  COVID-19 is a structural break to monitor.\n\n")
cat("  Proceed to Chapter 3 — ARIMAX Modelling\n")
cat("================================================\n")
```