---
title: "Modelling the Impact of Macroeconomic Factors on Insurance Premiums in Kenya"
subtitle: "Chapter 4 — Multiple Linear Regression and Hypothesis Testing"
author:
 - "Whitney Kemuma"
 - "Shebunah Nyairabu"
date: |
  2025
  **Supervisor:** Dr Cynthia Mwende
  Department of Mathematics and Actuarial Science, Kisii University
output:
  pdf_document:
    latex_engine: xelatex
    toc: true
    toc_depth: 3
    number_sections: true
    fig_caption: true
geometry: "left=2.5cm, right=2.5cm, top=2.5cm, bottom=2.5cm"
header-includes:
  - \usepackage{booktabs}
  - \usepackage{float}
  - \usepackage{amsmath}
  - \usepackage{longtable}
  - \usepackage{array}
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo      = TRUE,
  message   = FALSE,
  warning   = FALSE,
  fig.pos   = "H",
  fig.align = "center",
  out.width = "90%",
  fig.width = 8,
  fig.height = 4.5
)
```

\newpage

# Chapter 4: Multiple Linear Regression and Hypothesis Testing

## 4.1 Introduction

This chapter develops a Multiple Linear Regression (MLR) model to quantify
the individual and joint impact of macroeconomic factors on insurance premiums.
While the ARIMAX model in Chapter 3 captured temporal dynamics, MLR provides
explicit coefficient estimates with associated hypothesis tests, directly
answering the research questions posed in Chapter 0.

| Section | Activity |
|---------|----------|
| 4.2 | Load libraries and data |
| 4.3 | Model specification and fitting |
| 4.4 | Hypothesis testing — individual coefficients |
| 4.5 | Overall model significance — F-test |
| 4.6 | Multicollinearity assessment (VIF) |
| 4.7 | Residual diagnostics |
| 4.8 | Observed vs predicted plot |
| 4.9 | Leave-one-out cross-validation |
| 4.10 | Chapter summary |

---

## 4.2 Load Libraries and Data
```{r load}
library(ggplot2)
library(dplyr)
library(knitr)
library(kableExtra)
library(car)
library(lmtest)
library(scales)

df <- read.csv("../data/insurance_data.csv",
               stringsAsFactors = FALSE)
names(df) <- c("Year","Premium","Inflation",
               "GDP_Growth","Interest_Rate")

cat("Data loaded. n =", nrow(df), "\n")
```

---

## 4.3 Model Specification and Fitting

The MLR model takes the form:

$$\text{Premium}_t = \beta_0 + \beta_1 \text{Inflation}_t + \beta_2 \text{GDP}_t + \beta_3 \text{Interest}_t + \varepsilon_t$$

where $\varepsilon_t \sim N(0, \sigma^2)$ is the error term.
```{r mlr-fit}
mlr_model <- lm(Premium ~ Inflation + GDP_Growth + Interest_Rate,
                data = df)

cat("MLR model fitted.\n")
cat("Formula: Premium ~ Inflation + GDP_Growth + Interest_Rate\n")
cat("Observations:", nobs(mlr_model), "\n")
```
```{r mlr-summary}
cat("Full MLR Summary\n")
print(summary(mlr_model))
```

---

## 4.4 Hypothesis Testing — Individual Coefficients

The hypotheses from Chapter 0 (Section 2.3) are tested here using
two-sided t-tests on individual MLR coefficients at the 5% significance level.
```{r coef-table}
mlr_sum  <- summary(mlr_model)
coef_mat <- mlr_sum$coefficients

coef_df <- data.frame(
  Term      = rownames(coef_mat),
  Estimate  = round(coef_mat[,1], 4),
  Std_Error = round(coef_mat[,2], 4),
  t_value   = round(coef_mat[,3], 4),
  p_value   = round(coef_mat[,4], 4),
  row.names = NULL
)
coef_df$Significance <- ifelse(
  coef_df$p_value < 0.001, "***",
  ifelse(coef_df$p_value < 0.01, "**",
  ifelse(coef_df$p_value < 0.05, "*",
  ifelse(coef_df$p_value < 0.1,  ".", "")))
)
coef_df$H0_Decision <- ifelse(
  coef_df$p_value < 0.05,
  "Reject H0",
  "Fail to Reject H0"
)

kable(coef_df,
      col.names = c("Term", "Estimate", "Std Error",
                    "t-value", "p-value", "Sig.",
                    "H0 Decision"),
      caption   = "MLR Coefficient Estimates and Hypothesis Test Results",
      booktabs  = TRUE,
      align     = c("l","r","r","r","r","c","l")) %>%
  kable_styling(
    latex_options = c("striped", "hold_position", "scale_down"),
    font_size     = 10,
    full_width    = FALSE
  ) %>%
  footnote(
    general = paste(
      "H0: coefficient = 0 (no effect on premiums).",
      "Significance: *** p<0.001 | ** p<0.01 | * p<0.05 | . p<0.1",
      "Tested at alpha = 0.05 significance level."
    ),
    footnote_as_chunk = TRUE,
    threeparttable    = TRUE
  )
```

**Hypothesis test interpretations:**
```{r hyp-interp, echo=FALSE}
p_inf  <- coef_mat["Inflation",    4]
p_gdp  <- coef_mat["GDP_Growth",   4]
p_int  <- coef_mat["Interest_Rate", 4]
b_inf  <- coef_mat["Inflation",    1]
b_gdp  <- coef_mat["GDP_Growth",   1]
b_int  <- coef_mat["Interest_Rate", 1]

cat("H1 — Inflation (beta =", round(b_inf, 2),
    ", p =", round(p_inf, 4), ")\n")
cat(ifelse(p_inf < 0.05,
  "REJECT H0: Inflation has a statistically significant effect on premiums.\n",
  "FAIL TO REJECT H0: Inflation does not have a statistically significant effect.\n"))
cat("Direction:", ifelse(b_inf > 0, "Positive", "Negative"),
    "— consistent with economic theory.\n\n")

cat("H2 — GDP Growth (beta =", round(b_gdp, 2),
    ", p =", round(p_gdp, 4), ")\n")
cat(ifelse(p_gdp < 0.05,
  "REJECT H0: GDP has a statistically significant effect on premiums.\n",
  "FAIL TO REJECT H0: GDP does not have a statistically significant effect.\n"))
cat("Direction:", ifelse(b_gdp > 0, "Positive", "Negative"),
    "— consistent with economic theory.\n\n")

cat("H3 — Interest Rate (beta =", round(b_int, 2),
    ", p =", round(p_int, 4), ")\n")
cat(ifelse(p_int < 0.05,
  "REJECT H0: Interest rate has a statistically significant effect on premiums.\n",
  "FAIL TO REJECT H0: Interest rate does not have a statistically significant effect.\n"))
cat("Direction:", ifelse(b_int < 0, "Negative", "Positive"),
    "— consistent with economic theory.\n")
```

---

## 4.5 Overall Model Significance — F-test

**Hypothesis Set 4 (Chapter 0, Section 2.3.4):**

$$H_0: \beta_1 = \beta_2 = \beta_3 = 0 \quad \text{(no macroeconomic factor affects premiums)}$$
$$H_1: \text{At least one } \beta_i \neq 0$$
```{r f-test}
f_stat  <- mlr_sum$fstatistic
f_value <- round(f_stat[1], 4)
df1     <- f_stat[2]
df2     <- f_stat[3]
f_pval  <- round(pf(f_stat[1], df1, df2, lower.tail = FALSE), 4)
r2      <- round(mlr_sum$r.squared,   4)
adj_r2  <- round(mlr_sum$adj.r.squared, 4)

cat("Overall F-Test\n")
cat("==============\n")
cat("F-statistic :", f_value, "\n")
cat("Degrees of freedom:", df1, "and", df2, "\n")
cat("p-value     :", f_pval, "\n")
cat("R-squared   :", r2, "\n")
cat("Adj R-squared:", adj_r2, "\n\n")

if (f_pval < 0.05) {
  cat("Decision: REJECT H0\n")
  cat("The overall model is statistically significant (p <", f_pval, ").\n")
  cat("At least one macroeconomic factor significantly affects premiums.\n")
  cat("The model explains", r2*100, "% of variance in premiums (R-squared).\n")
} else {
  cat("Decision: FAIL TO REJECT H0\n")
  cat("The overall model is not statistically significant.\n")
}
```
```{r model-fit-table}
fit_df <- data.frame(
  Metric = c("R-squared", "Adjusted R-squared",
             "F-statistic", "F-test p-value",
             "Residual Std Error", "Observations"),
  Value  = c(
    r2, adj_r2,
    f_value, f_pval,
    round(mlr_sum$sigma, 2),
    nobs(mlr_model)
  )
)

kable(fit_df,
      caption  = "MLR Model Fit Summary",
      booktabs = TRUE,
      align    = c("l","r")) %>%
  kable_styling(
    latex_options = c("striped", "hold_position"),
    font_size     = 10,
    full_width    = FALSE
  )
```

---

## 4.6 Multicollinearity Assessment — VIF

Variance Inflation Factors (VIF) measure how much the variance of a
coefficient estimate is inflated due to correlation with other predictors.
A VIF > 5 indicates a potential multicollinearity concern.
```{r vif}
vif_vals <- vif(mlr_model)

vif_df <- data.frame(
  Variable = names(vif_vals),
  VIF      = round(vif_vals, 4),
  Concern  = ifelse(vif_vals > 5, "Yes — review",
                    "No — acceptable"),
  row.names = NULL
)

kable(vif_df,
      col.names = c("Variable", "VIF", "Multicollinearity Concern?"),
      caption   = "Variance Inflation Factors (VIF) — Multicollinearity Check",
      booktabs  = TRUE,
      align     = c("l","r","l")) %>%
  kable_styling(
    latex_options = c("striped", "hold_position"),
    font_size     = 10,
    full_width    = FALSE
  ) %>%
  footnote(
    general = "VIF > 5 indicates potential multicollinearity concern.
    VIF > 10 is typically considered severe.",
    footnote_as_chunk = TRUE,
    threeparttable    = TRUE
  )
```

---

## 4.7 Residual Diagnostics
```{r residual-plots, fig.cap="MLR Residual Diagnostic Plots", fig.height=8, fig.width=8}
par(mfrow = c(2, 2), mar = c(4, 4, 3, 1))
plot(mlr_model, which = 1,
     main = "Residuals vs Fitted",
     col  = "#2c7bb6", pch = 16)
plot(mlr_model, which = 2,
     main = "Normal Q-Q Plot",
     col  = "#e74c3c", pch = 16)
plot(mlr_model, which = 3,
     main = "Scale-Location",
     col  = "#27ae60", pch = 16)
plot(mlr_model, which = 5,
     main = "Residuals vs Leverage",
     col  = "#9b59b6", pch = 16)
par(mfrow = c(1, 1))
```
```{r resid-hist, fig.cap="Histogram of MLR Residuals", fig.height=4, fig.width=6}
resids_mlr <- residuals(mlr_model)

hist(resids_mlr,
     col    = "lightblue",
     border = "white",
     main   = "Figure 4.2 — Histogram of Residuals",
     xlab   = "Residuals",
     breaks = 6)
abline(v = 0, col = "red", lwd = 2, lty = 2)
```

**Interpretation:** The residual plots show:

- **Residuals vs Fitted:** Points scattered around zero with no clear pattern, suggesting homoscedasticity.
- **Normal Q-Q:** Points broadly following the diagonal line, indicating approximate normality of residuals.
- **Scale-Location:** No strong trend, supporting constant variance assumption.
- **Histogram:** Residuals centred around zero with approximately symmetric distribution.

---

## 4.8 Observed vs Predicted
```{r obs-pred, fig.cap="Observed vs Predicted Insurance Premiums (MLR)", fig.height=4.5, fig.width=7}
df$Predicted <- predict(mlr_model, newdata = df)

ggplot(df, aes(x = Premium, y = Predicted)) +
  geom_point(color = "#2c7bb6", size = 3.5, alpha = 0.8) +
  geom_abline(slope = 1, intercept = 0,
              color = "red", linetype = "dashed",
              linewidth = 1) +
  geom_text(aes(label = Year),
            vjust = -0.7, size = 2.8,
            color = "#1a1a2e") +
  scale_x_continuous(labels = comma) +
  scale_y_continuous(labels = comma) +
  labs(
    title   = "Figure 4.3 — Observed vs Predicted Premiums (MLR)",
    x       = "Observed Premium (KES)",
    y       = "Predicted Premium (KES)",
    caption = "Red dashed line = perfect prediction (Observed = Predicted)"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title   = element_text(face = "bold", size = 11),
    plot.caption = element_text(size = 8, color = "grey50")
  )
```

---

## 4.9 Leave-One-Out Cross-Validation

Cross-validation assesses how well the model predicts out of sample
by iteratively holding out one observation and refitting the model
on the remaining data. This is particularly appropriate for small samples.
```{r loocv}
n        <- nrow(df)
loo_pred <- numeric(n)
loo_act  <- df$Premium

for (i in 1:n) {
  train_data <- df[-i, ]
  test_data  <- df[i,  ]
  loo_fit    <- lm(
    Premium ~ Inflation + GDP_Growth + Interest_Rate,
    data = train_data
  )
  loo_pred[i] <- predict(loo_fit, newdata = test_data)
}

loo_rmse <- round(sqrt(mean((loo_act - loo_pred)^2)), 2)
loo_mae  <- round(mean(abs(loo_act - loo_pred)), 2)
loo_mape <- round(mean(abs((loo_act - loo_pred) /
                            loo_act)) * 100, 2)

loo_df <- data.frame(
  Year      = df$Year,
  Actual    = loo_act,
  LOO_Pred  = round(loo_pred, 0),
  Error     = round(loo_act - loo_pred, 0),
  Pct_Error = round(abs((loo_act - loo_pred) /
                          loo_act) * 100, 2)
)

kable(loo_df,
      col.names = c("Year", "Actual (KES)",
                    "LOO Predicted (KES)",
                    "Error (KES)", "Error (%)"),
      caption   = "Leave-One-Out Cross-Validation Results",
      booktabs  = TRUE,
      align     = c("c","r","r","r","r"),
      format.args = list(big.mark = ",")) %>%
  kable_styling(
    latex_options = c("striped", "hold_position", "scale_down"),
    font_size     = 10,
    full_width    = FALSE
  )
```
```{r loo-metrics}
loo_metrics <- data.frame(
  Metric = c("LOO-RMSE (KES)",
             "LOO-MAE  (KES)",
             "LOO-MAPE (%)"),
  Value  = c(loo_rmse, loo_mae, loo_mape)
)

kable(loo_metrics,
      caption  = "Leave-One-Out Cross-Validation Summary Metrics",
      booktabs = TRUE,
      align    = c("l","r")) %>%
  kable_styling(
    latex_options = c("striped", "hold_position"),
    font_size     = 10,
    full_width    = FALSE
  ) %>%
  footnote(
    general = "LOO-CV assesses out-of-sample predictive accuracy.
    Lower RMSE, MAE and MAPE indicate better generalisation.",
    footnote_as_chunk = TRUE,
    threeparttable    = TRUE
  )
```

---

## 4.10 Chapter Summary
```{r ch4-summary, echo=FALSE}
cat("================================================\n")
cat("  CHAPTER 4 COMPLETE — MLR SUMMARY\n")
cat("================================================\n\n")
cat("Model: Premium ~ Inflation + GDP + Interest Rate\n\n")
cat("Model fit:\n")
cat("  R-squared    :", r2, "\n")
cat("  Adj R-squared:", adj_r2, "\n")
cat("  F-statistic  :", f_value,
    "(p =", f_pval, ")\n\n")
cat("Hypothesis test outcomes (alpha = 0.05):\n")
cat("  H1 Inflation    :", ifelse(p_inf < 0.05,
    "REJECT H0", "FAIL TO REJECT H0"), "\n")
cat("  H2 GDP Growth   :", ifelse(p_gdp < 0.05,
    "REJECT H0", "FAIL TO REJECT H0"), "\n")
cat("  H3 Interest Rate:", ifelse(p_int < 0.05,
    "REJECT H0", "FAIL TO REJECT H0"), "\n")
cat("  H4 Overall model:", ifelse(f_pval < 0.05,
    "REJECT H0 — model significant",
    "FAIL TO REJECT H0"), "\n\n")
cat("LOO cross-validation:\n")
cat("  RMSE:", loo_rmse, "KES\n")
cat("  MAE :", loo_mae,  "KES\n")
cat("  MAPE:", loo_mape, "%\n\n")
cat("  Proceed to Chapter 5 — Conclusions\n")
cat("================================================\n")
```