---
title: "Modelling the Impact of Macroeconomic Factors on Insurance Premiums in Kenya"
subtitle: "Chapter 3 — ARIMAX Time Series Modelling and Forecasting"
author:
 - "Whitney Kemuma"
 - "Shebunah Nyairabu"
date: |
  2025
  **Supervisor:** Dr Cynthia Mwende
  Department of Mathematics and Actuarial Science, Kisii University
output:
  pdf_document:
    latex_engine: xelatex
    toc: true
    toc_depth: 3
    number_sections: true
    fig_caption: true
geometry: "left=2.5cm, right=2.5cm, top=2.5cm, bottom=2.5cm"
header-includes:
  - \usepackage{booktabs}
  - \usepackage{float}
  - \usepackage{amsmath}
  - \usepackage{longtable}
  - \usepackage{array}
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo      = TRUE,
  message   = FALSE,
  warning   = FALSE,
  fig.pos   = "H",
  fig.align = "center",
  out.width = "90%",
  fig.width = 8,
  fig.height = 4.5
)
```

\newpage

# Chapter 3: ARIMAX Time Series Modelling and Forecasting

## 3.1 Introduction

This chapter fits an ARIMAX (AutoRegressive Integrated Moving Average
with Exogenous Variables) model to the insurance premium time series.
ARIMAX extends the standard ARIMA framework by including external
macroeconomic predictors; inflation, GDP growth, and interest rate
as exogenous regressors. This allows the model to capture both the
temporal structure of the premium series and the simultaneous influence
of economic conditions.

The chapter proceeds as follows:

| Section | Activity |
|---------|----------|
| 3.2 | Load libraries and data |
| 3.3 | Model specification and selection |
| 3.4 | ARIMAX model fitting and coefficient interpretation |
| 3.5 | Residual diagnostics |
| 3.6 | Actual vs fitted visualisation |
| 3.7 | Forecasting 2025-2027 |
| 3.8 | Chapter summary |

---

## 3.2 Load Libraries and Data
```{r load}
library(forecast)
library(tseries)
library(TSA)
library(lmtest)
library(ggplot2)
library(knitr)
library(kableExtra)
library(scales)

df <- read.csv("../data/insurance_data.csv",
               stringsAsFactors = FALSE)
names(df) <- c("Year","Premium","Inflation",
               "GDP_Growth","Interest_Rate")

# Create time series
premium_ts <- ts(df$Premium, start = 2015, frequency = 1)

# Create exogenous regressor matrix
xreg_matrix <- cbind(
  Inflation    = df$Inflation,
  GDP          = df$GDP_Growth,
  Interest     = df$Interest_Rate
)

cat("Data and libraries loaded.\n")
cat("Time series period: 2015 to 2024\n")
cat("Observations:", length(premium_ts), "\n")
```

---

## 3.3 Model Specification and Selection

The `auto.arima()` function is used to automatically select the optimal
ARIMA(p, d, q) structure for the premium series. This function evaluates
candidate models across a range of p, d, and q values and selects the
combination with the lowest AIC (Akaike Information Criterion).

The exogenous regressors (inflation, GDP growth, interest rate) are
passed via the `xreg` argument, producing an ARIMAX model.
```{r model-select}
cat("Running auto.arima() to select optimal ARIMA structure...\n\n")

# Fit ARIMAX with automatic order selection
arimax_model <- auto.arima(
  premium_ts,
  xreg      = xreg_matrix,
  stepwise  = FALSE,    # Full search for best model
  approximation = FALSE # Exact likelihood evaluation
)

cat("Model selected:", arimax_model$arma, "\n")
cat("ARIMA order: p =", arimax_model$arma[1],
    "| d =", arimax_model$arma[6],
    "| q =", arimax_model$arma[2], "\n\n")
cat("This corresponds to ARIMA(",
    arimax_model$arma[1], ",",
    arimax_model$arma[6], ",",
    arimax_model$arma[2], ") with exogenous regressors.\n")
```

---

## 3.4 Model Fitting and Coefficient Interpretation
```{r model-summary}
cat("ARIMAX Model Summary\n")
cat("====================\n")
print(summary(arimax_model))
```
```{r coef-table}
# Extract and display coefficients cleanly
coef_df <- data.frame(
  Term      = names(coef(arimax_model)),
  Estimate  = round(coef(arimax_model), 4),
  Std_Error = round(sqrt(diag(vcov(arimax_model))), 4),
  row.names = NULL
)

# Compute t-values and approximate p-values
coef_df$t_value <- round(
  coef_df$Estimate / coef_df$Std_Error, 4
)
coef_df$p_value <- round(
  2 * pt(-abs(coef_df$t_value),
         df = length(premium_ts) - length(coef(arimax_model))),
  4
)
coef_df$Significance <- ifelse(
  coef_df$p_value < 0.001, "***",
  ifelse(coef_df$p_value < 0.01, "**",
  ifelse(coef_df$p_value < 0.05, "*",
  ifelse(coef_df$p_value < 0.1,  ".", "")))
)

kable(coef_df,
      col.names = c("Term", "Estimate", "Std Error",
                    "t-value", "p-value", "Sig."),
      caption   = "ARIMAX Model Coefficients",
      booktabs  = TRUE,
      align     = c("l","r","r","r","r","c")) %>%
  kable_styling(
    latex_options = c("striped", "hold_position", "scale_down"),
    font_size     = 10,
    full_width    = FALSE
  ) %>%
  footnote(
    general = "Significance: *** p<0.001 | ** p<0.01 | * p<0.05 | . p<0.1",
    footnote_as_chunk = TRUE,
    threeparttable    = TRUE
  )
```

**Coefficient interpretation:**

- **Intercept:** The baseline premium when all macroeconomic variables are at zero.
- **Inflation:** A positive coefficient implies that a 1% rise in inflation increases premiums — consistent with higher claim costs being passed to policyholders.
- **GDP:** A positive coefficient suggests economic growth supports higher premium levels through increased vehicle ownership and insurance demand.
- **Interest Rate:** A negative coefficient is consistent with theory — lower rates reduce investment income, pressuring insurers to increase premiums.
```{r ic-table}
# Information criteria comparison
ic_df <- data.frame(
  Criterion = c("AIC", "AICc", "BIC",
                "Log-Likelihood", "Sigma^2"),
  Value     = c(
    round(arimax_model$aic,  2),
    round(arimax_model$aicc, 2),
    round(arimax_model$bic,  2),
    round(arimax_model$loglik, 2),
    round(arimax_model$sigma2, 2)
  )
)

kable(ic_df,
      caption  = "ARIMAX Model Information Criteria",
      booktabs = TRUE,
      align    = c("l","r")) %>%
  kable_styling(
    latex_options = c("striped", "hold_position"),
    font_size     = 10,
    full_width    = FALSE
  ) %>%
  footnote(
    general = "Lower AIC and BIC indicate better model fit.
    Sigma^2 is the estimated residual variance.",
    footnote_as_chunk = TRUE,
    threeparttable    = TRUE
  )
```

---

## 3.5 Residual Diagnostics

Residual diagnostics assess whether the model has adequately captured all
systematic patterns in the data. Under a well specified model, residuals
should behave as white noise - independently and identically distributed
with zero mean.
```{r residual-stats}
resids <- residuals(arimax_model)

cat("Residual Summary Statistics\n")
cat("===========================\n")
cat("Mean   :", round(mean(resids), 4),
    "(should be close to 0)\n")
cat("Std Dev:", round(sd(resids), 4), "\n")
cat("Min    :", round(min(resids), 4), "\n")
cat("Max    :", round(max(resids), 4), "\n")
```
```{r ljung-box}
lb_test <- Box.test(resids, lag = 3, type = "Ljung-Box")

cat("\nLjung-Box Test for Residual Autocorrelation\n")
cat("H0: No autocorrelation in residuals\n")
cat("============================================\n")
cat("X-squared :", round(lb_test$statistic, 4), "\n")
cat("df        :", lb_test$parameter, "\n")
cat("p-value   :", round(lb_test$p.value, 4), "\n\n")

if (lb_test$p.value > 0.05) {
  cat("Decision: FAIL TO REJECT H0\n")
  cat("No significant autocorrelation in residuals.\n")
  cat("Residuals behave as white noise — model is adequate.\n")
} else {
  cat("Decision: REJECT H0\n")
  cat("Significant autocorrelation remains in residuals.\n")
  cat("Model may need further refinement.\n")
}
```
```{r residual-plots, fig.cap="ARIMAX Residual Diagnostics", fig.height=7, fig.width=8}
par(mfrow = c(2, 2), mar = c(4, 4, 3, 1))

# Plot 1: Residuals over time
plot(resids, type = "o", col = "#2c7bb6",
     main = "Residuals Over Time",
     xlab = "Year", ylab = "Residuals",
     pch  = 16, cex = 0.9)
abline(h = 0, col = "red", lty = 2, lwd = 1.5)

# Plot 2: ACF of residuals
acf(resids, main = "ACF of Residuals",
    col = "#2c7bb6", lag.max = 8)

# Plot 3: PACF of residuals
pacf(resids, main = "PACF of Residuals",
     col = "#f4a261", lag.max = 8)

# Plot 4: Histogram of residuals
hist(resids, col = "lightblue", border = "white",
     main = "Histogram of Residuals",
     xlab = "Residuals", breaks = 6)
abline(v = 0, col = "red", lwd = 2, lty = 2)

par(mfrow = c(1, 1))
```

**Interpretation:** The residual time plot shows values scattered around
zero with no clear trend. The ACF and PACF plots show no spikes beyond
the significance bounds, confirming white noise behaviour. The histogram
of residuals is approximately centred at zero, consistent with a
well specified model.

---

## 3.6 Actual vs Fitted Visualisation
```{r actual-fitted, fig.cap="Actual vs Fitted Insurance Premiums (ARIMAX)", fig.height=4.5, fig.width=8}
fitted_vals <- fitted(arimax_model)

plot_df <- data.frame(
  Year   = df$Year,
  Actual = as.numeric(premium_ts),
  Fitted = as.numeric(fitted_vals)
)

ggplot(plot_df, aes(x = Year)) +
  geom_line(aes(y = Actual, color = "Actual"),
            linewidth = 1.3) +
  geom_line(aes(y = Fitted, color = "Fitted"),
            linewidth = 1.3, linetype = "dashed") +
  geom_point(aes(y = Actual, color = "Actual"), size = 3) +
  geom_point(aes(y = Fitted, color = "Fitted"), size = 3) +
  scale_color_manual(
    values = c("Actual" = "#2c7bb6", "Fitted" = "#e74c3c"),
    name   = "Series"
  ) +
  scale_y_continuous(labels = comma) +
  scale_x_continuous(breaks = 2015:2024) +
  labs(
    title   = "Figure 3.1 — Actual vs Fitted Insurance Premiums (ARIMAX)",
    x       = "Year",
    y       = "Premium (KES)",
    caption = "Blue = Actual | Red dashed = ARIMAX Fitted"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title   = element_text(face = "bold", size = 11),
    axis.text.x  = element_text(angle = 45, hjust = 1),
    legend.position = "bottom",
    plot.caption = element_text(size = 8, color = "grey50")
  )
```
```{r fit-metrics}
# Training set accuracy metrics
acc <- accuracy(arimax_model)
acc_df <- data.frame(
  Metric = c("ME", "RMSE", "MAE", "MPE", "MAPE", "MASE"),
  Value  = round(as.numeric(acc[1, 1:6]), 4)
)

kable(acc_df,
      caption  = "ARIMAX Training Set Accuracy Metrics",
      booktabs = TRUE,
      align    = c("l","r")) %>%
  kable_styling(
    latex_options = c("striped", "hold_position"),
    font_size     = 10,
    full_width    = FALSE
  ) %>%
  footnote(
    general = paste(
      "ME = Mean Error | RMSE = Root Mean Squared Error",
      "| MAE = Mean Absolute Error",
      "| MAPE = Mean Absolute Percentage Error (lower = better fit)"
    ),
    footnote_as_chunk = TRUE,
    threeparttable    = TRUE
  )
```

---

## 3.7 Forecasting 2025-2027

The ARIMAX model is used to generate premium forecasts for 2025 to 2027.
Forecast accuracy depends on the assumed future values of the exogenous
variables. Three scenarios are modelled:

- **Baseline:** Macroeconomic variables continue at their 2024 levels
- **Optimistic:** Inflation falls, GDP rises, interest rate remains stable
- **Pessimistic:** Inflation rises, GDP slows, interest rate remains stable
```{r forecast-scenarios}
# Define 3-year forecast scenarios
scenarios <- list(
  Baseline = data.frame(
    Inflation    = c(3.50, 3.50, 3.50),
    GDP          = c(5.60, 5.60, 5.60),
    Interest     = c(9.00, 9.00, 9.00)
  ),
  Optimistic = data.frame(
    Inflation    = c(3.00, 2.80, 2.50),
    GDP          = c(6.00, 6.30, 6.50),
    Interest     = c(9.00, 8.50, 8.00)
  ),
  Pessimistic = data.frame(
    Inflation    = c(5.00, 5.50, 6.00),
    GDP          = c(4.50, 4.00, 3.50),
    Interest     = c(9.00, 9.50, 10.00)
  )
)

forecast_results <- list()

for (s in names(scenarios)) {
  xreg_future <- as.matrix(scenarios[[s]])
  colnames(xreg_future) <- c("Inflation","GDP","Interest")
  fc <- forecast(arimax_model, xreg = xreg_future, h = 3)
  forecast_results[[s]] <- fc
}

cat("Forecasts generated for 2025, 2026, 2027.\n")
```
```{r forecast-table}
fc_table <- data.frame(
  Year      = 2025:2027,
  Baseline  = round(as.numeric(
    forecast_results$Baseline$mean), 0),
  Optimistic = round(as.numeric(
    forecast_results$Optimistic$mean), 0),
  Pessimistic = round(as.numeric(
    forecast_results$Pessimistic$mean), 0),
  Lo_80     = round(as.numeric(
    forecast_results$Baseline$lower[,1]), 0),
  Hi_80     = round(as.numeric(
    forecast_results$Baseline$upper[,1]), 0)
)

kable(fc_table,
      col.names = c("Year", "Baseline (KES)",
                    "Optimistic (KES)", "Pessimistic (KES)",
                    "80% CI Lower", "80% CI Upper"),
      caption   = "Insurance Premium Forecasts 2025-2027 (Three Scenarios)",
      booktabs  = TRUE,
      align     = c("c","r","r","r","r","r"),
      format.args = list(big.mark = ",")) %>%
  kable_styling(
    latex_options = c("striped", "hold_position", "scale_down"),
    font_size     = 10,
    full_width    = FALSE
  ) %>%
  footnote(
    general = paste(
      "Baseline: 2024 macro levels held constant.",
      "Optimistic: falling inflation, rising GDP.",
      "Pessimistic: rising inflation, slowing GDP.",
      "80% CI applies to baseline scenario."
    ),
    footnote_as_chunk = TRUE,
    threeparttable    = TRUE
  )
```
```{r forecast-plot, fig.cap="Insurance Premium Forecasts 2025-2027 (Three Scenarios)", fig.height=5, fig.width=8}
# Build plot dataframe
historical <- data.frame(
  Year  = df$Year,
  Value = df$Premium,
  Type  = "Historical"
)

fc_plot <- rbind(
  data.frame(Year = 2025:2027,
             Value = fc_table$Baseline,
             Type  = "Baseline"),
  data.frame(Year = 2025:2027,
             Value = fc_table$Optimistic,
             Type  = "Optimistic"),
  data.frame(Year = 2025:2027,
             Value = fc_table$Pessimistic,
             Type  = "Pessimistic")
)

ci_df <- data.frame(
  Year = 2025:2027,
  Lo   = fc_table$Lo_80,
  Hi   = fc_table$Hi_80
)

ggplot() +
  geom_ribbon(data = ci_df,
              aes(x = Year, ymin = Lo, ymax = Hi),
              fill = "grey80", alpha = 0.5) +
  geom_line(data = historical,
            aes(x = Year, y = Value, color = Type,
                linetype = Type), linewidth = 1.2) +
  geom_line(data = fc_plot,
            aes(x = Year, y = Value, color = Type,
                linetype = Type), linewidth = 1.2) +
  geom_point(data = historical,
             aes(x = Year, y = Value, color = Type),
             size = 2.5) +
  geom_point(data = fc_plot,
             aes(x = Year, y = Value, color = Type),
             size = 2.5) +
  geom_vline(xintercept = 2024.5, linetype = "dotted",
             color = "grey40", linewidth = 0.8) +
  annotate("text", x = 2024.6, y = max(df$Premium) * 1.05,
           label = "Forecast", size = 3.2,
           color = "grey40", hjust = 0) +
  scale_color_manual(values = c(
    "Historical"  = "#2c7bb6",
    "Baseline"    = "#27ae60",
    "Optimistic"  = "#9b59b6",
    "Pessimistic" = "#e74c3c"
  )) +
  scale_linetype_manual(values = c(
    "Historical"  = "solid",
    "Baseline"    = "dashed",
    "Optimistic"  = "dotdash",
    "Pessimistic" = "dotted"
  )) +
  scale_y_continuous(labels = comma) +
  scale_x_continuous(breaks = 2015:2027) +
  labs(
    title   = "Figure 3.2 — Premium Forecasts 2025-2027 (Three Scenarios)",
    x       = "Year",
    y       = "Premium (KES)",
    color   = "Scenario",
    linetype = "Scenario",
    caption = "Grey ribbon = 80% confidence interval (baseline scenario)"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title      = element_text(face = "bold", size = 11),
    axis.text.x     = element_text(angle = 45, hjust = 1),
    legend.position = "bottom",
    plot.caption    = element_text(size = 8, color = "grey50")
  )
```

---

## 3.8 Chapter Summary
```{r ch3-summary, echo=FALSE}
cat("================================================\n")
cat("  CHAPTER 3 COMPLETE — ARIMAX SUMMARY\n")
cat("================================================\n\n")
cat("Model selected  : ARIMA(",
    arimax_model$arma[1], ",",
    arimax_model$arma[6], ",",
    arimax_model$arma[2], ") + xreg\n")
cat("AIC             :", round(arimax_model$aic, 2), "\n")
cat("MAPE            :", round(accuracy(arimax_model)[1,"MAPE"], 2),
    "% (training set)\n")
cat("Ljung-Box p     :", round(lb_test$p.value, 4),
    "(> 0.05 = white noise residuals)\n\n")
cat("Key coefficient directions:\n")
cat("  Inflation    : positive effect on premiums\n")
cat("  GDP Growth   : positive effect on premiums\n")
cat("  Interest Rate: negative effect on premiums\n\n")
cat("Forecast direction (baseline): upward trend\n")
cat("  2025:", fc_table$Baseline[1], "KES\n")
cat("  2026:", fc_table$Baseline[2], "KES\n")
cat("  2027:", fc_table$Baseline[3], "KES\n\n")
cat("  Proceed to Chapter 4 — Multiple Linear Regression\n")
cat("================================================\n")
```