---
title: "Modelling the Impact of Macroeconomic Factors on Insurance Premiums in Kenya"
subtitle: "Chapter 1 - Data"
author:
 - "Whitney Kemuma"
 - "Shebunah Nyairabu"
supervisor: "Dr Cynthia Mwende"
date: |
  2025
  **Supervisor:** Dr Cynthia Mwende
  Department of Mathematics and Actuarial Science, Kisii University
output:
  pdf_document:
    latex_engine: xelatex
    toc: true
    toc_depth: 3
    number_sections: true
    fig_caption: true
header-includes:
  - \usepackage{booktabs}
  - \usepackage{float}
  - \usepackage{amsmath}
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.pos = "H"
)
```

\newpage

# Chapter 1: Data Loading, Cleaning and Descriptive Statistics

## 1.1 Introduction

This chapter covers the full data preparation pipeline for this study.
The dataset consists of annual observations from 2015 to 2024 covering
four variables: insurance premiums, inflation rate, GDP growth rate, and
interest rate. All data was sourced from the Insurance Regulatory Authority
(IRA), Kenya National Bureau of Statistics (KNBS), and the Central Bank
of Kenya (CBK), as documented in Chapter 0.

The steps in this chapter are:

| Step | Activity |
|------|----------|
| 1.2 | Load required libraries |
| 1.3 | Load and inspect the dataset |
| 1.4 | Data cleaning and validation |
| 1.5 | Descriptive statistics |
| 1.6 | Variable definitions and measurement |

---

## 1.2 Load Required Libraries
```{r libraries}
# Core data manipulation and visualisation
library(ggplot2)      # Data visualisation
library(dplyr)        # Data manipulation
library(knitr)        # Table formatting
library(kableExtra)   # Enhanced table styling

# Time series packages
library(forecast)     # ARIMAX modelling and forecasting
library(tseries)      # Stationarity testing (ADF)
library(TSA)          # Time series tools
library(lmtest)       # Ljung-Box and Granger causality tests

# Regression diagnostics
library(car)          # VIF multicollinearity test
library(corrplot)     # Correlation matrix plots

cat("All libraries loaded successfully.\n")
```

---

## 1.3 Load and Inspect the Dataset
```{r load-data}
# Load the dataset from the data/ folder
# The path goes up one level from scripts/ into the project root
# then into data/
df <- read.csv("../data/insurance_data.csv", stringsAsFactors = FALSE)

# Display raw structure
cat("Dataset dimensions:", nrow(df), "rows x", ncol(df), "columns\n\n")
cat("Column names (raw):\n")
print(names(df))
```
```{r clean-names}
# Clean column names — remove whitespace and special characters
# from the raw CSV headers
names(df) <- c(
  "Year",
  "Premium",
  "Inflation",
  "GDP_Growth",
  "Interest_Rate"
)

cat("Cleaned column names:\n")
print(names(df))
cat("\nFirst 5 rows:\n")
print(head(df, 5))
```
```{r data-types}
# Check data types
cat("Data types per column:\n")
str(df)
```

---

## 1.4 Data Cleaning and Validation
```{r missing-values}
# Check for missing values
missing_count <- colSums(is.na(df))
cat("Missing values per column:\n")
print(missing_count)

if (sum(missing_count) == 0) {
  cat("\nNo missing values found. Dataset is complete.\n")
} else {
  cat("\nWarning: Missing values detected. Review before modelling.\n")
}
```
```{r duplicates}
# Check for duplicate rows
dup_count <- sum(duplicated(df))
cat("Duplicate rows:", dup_count, "\n")

if (dup_count == 0) {
  cat("No duplicates found.\n")
}
```
```{r range-validation}
# Validate numeric ranges — flag implausible values
cat("Variable ranges:\n")
cat("  Year          :", min(df$Year), "—", max(df$Year), "\n")
cat("  Premium (KES) :", min(df$Premium), "—", max(df$Premium), "\n")
cat("  Inflation (%)  :", min(df$Inflation), "—", max(df$Inflation), "\n")
cat("  GDP Growth (%) :", min(df$GDP_Growth), "—", max(df$GDP_Growth), "\n")
cat("  Interest Rate (%):", min(df$Interest_Rate), "—", max(df$Interest_Rate), "\n")

# Flag the 2020 negative GDP as expected (COVID-19)
cat("\nNote: GDP Growth of", df$GDP_Growth[df$Year == 2020],
    "in 2020 is consistent with COVID-19 economic contraction.\n")
```
```{r full-dataset-table}
# Display the complete cleaned dataset as a formatted table
kable(df,
      col.names = c("Year", "Premium (KES)", "Inflation (%)",
                    "GDP Growth (%)", "Interest Rate (%)"),
      caption   = "Complete Dataset: Kenya Insurance Premiums and
                   Macroeconomic Indicators (2015-2024)",
      booktabs  = TRUE,
      align     = c("c","r","r","r","r")) %>%
  kable_styling(
    latex_options = c("striped", "hold_position"),
    font_size     = 11
  ) %>%
  row_spec(6, bold = TRUE, color = "red",
           hline_after = FALSE) %>%  # Highlight 2020 COVID year
  footnote(
    general = "Red row (2020) marks the COVID-19 economic contraction year.
               Source: IRA, KNBS, CBK (2015-2024).",
    footnote_as_chunk = TRUE
  )
```

---

## 1.5 Descriptive Statistics
```{r descriptive-stats}
# Compute descriptive statistics for all numeric variables
desc_stats <- data.frame(
  Variable = c("Premium (KES)", "Inflation (%)",
               "GDP Growth (%)", "Interest Rate (%)"),
  N        = rep(nrow(df), 4),
  Mean     = c(mean(df$Premium),      mean(df$Inflation),
               mean(df$GDP_Growth),   mean(df$Interest_Rate)),
  Median   = c(median(df$Premium),    median(df$Inflation),
               median(df$GDP_Growth), median(df$Interest_Rate)),
  SD       = c(sd(df$Premium),        sd(df$Inflation),
               sd(df$GDP_Growth),     sd(df$Interest_Rate)),
  Min      = c(min(df$Premium),       min(df$Inflation),
               min(df$GDP_Growth),    min(df$Interest_Rate)),
  Max      = c(max(df$Premium),       max(df$Inflation),
               max(df$GDP_Growth),    max(df$Interest_Rate))
)

# Round numeric columns
desc_stats[, 3:7] <- round(desc_stats[, 3:7], 3)

kable(desc_stats,
      caption  = "Descriptive Statistics: All Variables (2015-2024)",
      booktabs = TRUE,
      align    = c("l","c","r","r","r","r","r")) %>%
  kable_styling(
    latex_options = c("striped", "hold_position"),
    font_size     = 10
  )
```
```{r interpret-desc, echo=FALSE}
cat("Key observations from descriptive statistics:\n\n")

cat("Premium:\n")
cat("  Mean KES", round(mean(df$Premium), 0),
    "| Range KES", min(df$Premium), "to KES", max(df$Premium), "\n")
cat("  Premiums rose by", round((max(df$Premium) -
    min(df$Premium)) / min(df$Premium) * 100, 1),
    "% over the study period.\n\n")

cat("Inflation:\n")
cat("  Mean", round(mean(df$Inflation), 2),
    "% | SD", round(sd(df$Inflation), 2), "%\n")
cat("  Peaked at", max(df$Inflation), "% in",
    df$Year[which.max(df$Inflation)],
    "and dropped to", min(df$Inflation), "% in",
    df$Year[which.min(df$Inflation)], "\n\n")

cat("GDP Growth:\n")
cat("  Mean", round(mean(df$GDP_Growth), 2), "%\n")
cat("  Only negative value:", min(df$GDP_Growth),
    "% in", df$Year[which.min(df$GDP_Growth)],
    "(COVID-19 contraction)\n\n")

cat("Interest Rate:\n")
cat("  Declined monotonically from", max(df$Interest_Rate),
    "% in", df$Year[which.max(df$Interest_Rate)],
    "to", min(df$Interest_Rate), "% in",
    df$Year[which.min(df$Interest_Rate)], "\n")
cat("  Total decline:", max(df$Interest_Rate) -
    min(df$Interest_Rate), "percentage points\n")
```

---

## 1.6 Variable Definitions and Measurement
```{r variable-table, echo=FALSE}
var_table <- data.frame(
  Variable = c(
    "Insurance Premium (KES)",
    "Inflation Rate (%)",
    "GDP Growth Rate (%)",
    "Interest Rate (%)"
  ),
  Role = c(
    "Dependent variable",
    "Independent variable",
    "Independent variable",
    "Independent variable"
  ),
  Definition = c(
    "Annual comprehensive car insurance premium for a vehicle valued at KES 2,000,000",
    "Annual consumer price inflation rate as published by KNBS",
    "Annual real GDP growth rate as published by the World Bank and KNBS",
    "CBK benchmark lending rate used as a proxy for the prevailing interest rate environment"
  ),
  Expected_Effect = c(
    "—",
    "Positive: higher inflation increases claim costs and premiums",
    "Positive: economic growth increases vehicle ownership and insurance demand",
    "Negative: lower rates reduce investment income, pressuring premiums upward"
  )
)

kable(var_table,
      col.names = c("Variable", "Role", "Definition", "Expected Effect"),
      caption   = "Variable Definitions, Roles and Expected Effects",
      booktabs  = TRUE) %>%
  kable_styling(
    latex_options = c("striped", "hold_position", "scale_down"),
    font_size     = 9
  ) %>%
  column_spec(3, width = "6cm") %>%
  column_spec(4, width = "5cm")
```

---

## 1.7 Chapter Summary
```{r chapter-summary, echo=FALSE}
cat("========================================================\n")
cat("  CHAPTER 1 COMPLETE — DATA PREPARATION SUMMARY\n")
cat("========================================================\n")
cat("  Observations          :", nrow(df), "(2015-2024)\n")
cat("  Variables             : 4 (1 dependent, 3 independent)\n")
cat("  Missing values        : 0\n")
cat("  Duplicate rows        : 0\n")
cat("  COVID-19 flag (2020)  : GDP growth =",
    df$GDP_Growth[df$Year == 2020], "%\n")
cat("  Premium growth        :", round((max(df$Premium) -
    min(df$Premium)) / min(df$Premium) * 100, 1), "% over study period\n")
cat("  Interest rate decline :", max(df$Interest_Rate) -
    min(df$Interest_Rate), "pp over study period\n")
cat("========================================================\n")
cat("  Proceed to Chapter 2 — Exploratory Data Analysis\n")
cat("========================================================\n")
```